Adapting the Chumbley Score to match striae on Land Engraved Areas of bullets

### 2018-08-06 jfs_submission/paper.Rmd
getting ready for review

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- land engraved areas (LEAs)
- algorithm
- signatures 
- same-source problem
bibliography: bibliography
abstract: The same-source problem remains a major challenge in forensic toolmark and firearm examination. Technological advances in surface imaging allow measurements of 3D surfaces at previously unforeseen resolutions and enable digitized imaging. Here, we investigate the applicability of the Chumbley scoring method \citep{hadler}, developed for screwdriver markings, for same-source identification of bullet striae. We provide methods to identify parameters that minimize the error rates for matching of LEAs using Hamby datasets 44 and 252 measured by NIST and CSAFE. We suggest a remedial algorithm to alleviate the problem of failed tests in the method, increasing both the power of the test and reducing error rates. Type II error rates of the proposed method improve by more than one third (Type I error of 0.05) and are on average about 0.22. This puts the proposed method on similar footing as other single feature matching approaches in the literature. 
---

@@

We then define the pairs for different-shifts as
\begin{equation}\label{eq.diff2}
\left(s_{(j)}^{(1)}, s_{(I-j+1)}^{(1)} \right) \text{ for } j = 
\begin{cases}
1, ..., I & \text{ for even } I \\
1, ..., (I-1)/2, (I-1)/2 + 2, ..., I & \text{ for odd } I
\end{cases},
\end{equation}
i.e.\ for an odd number of same-shift correlations, we skip the middle pair for the different-shift correlations (see \autoref{sketch-diff-2}).
This pairing ensures that the number of different-shift pairings is the same or at most one less than the number of same-shift pairings in all tests.
In the remainder of the paper, we will refer to the algorithm defined by \citet{hadler} as \textbf{(CS1)} and the suggested modified algorithm as \textbf{(CS2)} and compare their performance on the available scans of the Hamby study.
\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-diff-2.png}
\caption{\label{sketch-diff-2}Sketch of adjusted different-shift pairings. At most one of the same-shift pairings can not be matched with a different-shift pair under algorithm (CS2). }
\end{figure}

@@

errorsCV <- rbind(errorsCV1, errorsCV2)
ggerrors <- errorsCV %>%
  ggplot(aes(x = wo, y = beta, colour=factor(alpha))) +
  geom_point(aes(shape=factor(wv)), size=2.5) +
  geom_smooth(aes(group=alpha), se=FALSE, method="loess", span=1,  size=.7) +
  theme_bw() +
  scale_colour_brewer(expression("Nominal\ntype I error"~alpha),
                      palette="Set2") +
  scale_shape_discrete(expression("Size of\nvalidation window "~w[v])) +
  xlab(expression("Window size for optimization "~w[o])) +
  ylab("Type II error rate") +
  facet_wrap(~method, labeller="label_both")

@@

The results from Figures \ref{fig:roc} and \ref{fig:eer} are summarized in numbers in \autoref{tab:aucs}. Equal error rates (EER), rates for false positives (FPR) and false negatives (FNR) are shown side by side with the area under the curve (AUR) for both methods for a set of different optimization windows $w_o$ and a validation window $w_v$ of 30 pixels. The rate of false positive same-source identifications is equal to the statistical type I error rate, which is set to $\alpha = 0.05$ for this example. The rate of false negatives are missed same-source markings. This rate is also known as the type II error rate.   
Area under the curve (AUC) is shown with confidence intervals as given by \citet{delong}. CS2 significantly outperforms CS1 with respect to its predictive power in most situations.

@@

cvcs %>% filter(!wo %in% c(130, 140)) %>%
  ggplot(aes(d = as.numeric(match), m = 1-p_value)) + 
  geom_roc(labels=FALSE, n.cuts=500, pointsize=0, size=0.6, 
           aes(colour = method.wo)) +
  facet_wrap(~wv, labeller = label_bquote(w[v] == .(wv))) +
  scale_alpha_discrete(range=c(0.1, 0.6)) +
  scale_colour_manual(expression("Method /\nOptimization\nwindow size"~w[o]), values=cols) +
  theme_bw() +
  xlab("False Positives Rate (1 - Specificity)") +
  ylab("True Positives Rate (Sensitivity)") +
  guides(
   colour = guide_legend(ncol=2)
  ) +
  geom_abline(slope=-1, intercept = 1, linetype = 3, 
              colour="grey30", size = 0.25)

>>>>

---
title: "Adapting the Chumbley Score to match striae on Land Engraved Areas of bullets"
authors:
# - affiliation: Department of Statistics, Iowa State University 
- name: Ganesh Krishnan,
  degree: Master of Science (M.Sc.)
  # thank: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. We would also like to thank the efforts of CSAFE and Roy J carver Microscopy lab in scanning the Hamby-44 set and providing the scans to us.
# - affiliation: Department of Statistics and CSAFE, Iowa State University 
- name: Prof. Heike Hofmann,
  degree: Doctor of Philosophy (Ph.D.)
affiliation: Department of Statistics, Iowa State University and CSAFE (Center for Statistics and Applications in Forensic Evidence)
funding: The authors gratefully acknowledge funding from CSAFE through cooperative agreement \# 70NANB15H176 between NIST and Iowa State University. 
thank: We would like to thank the efforts of CSAFE and Roy J Carver Microscopy lab in scanning the Hamby-44 set and providing the scans to us. We would also like to thank NIST for providing the Hamby-44 bullets to us for scanning.
presentedat:  This work was presented in the form of a poster in the AFTE 2018 49th Annual Training Seminar, held from June 3rd to June 8th, 2018 in Charleston, West Virginia
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- forensic science
- toolmark
- cross-correlation
- Mann-Whitney U statistic
- land engraved areas (LEAs)
- algorithm
- signatures 
- same-source problem
bibliography: bibliography
abstract: The same-source problem remains a major challenge in forensic toolmark and firearm examination. Technological advances in surface imaging allow measurements of 3D surfaces at previously unforeseen resolutions and enable digitized imaging. Here, we investigate the applicability of the Chumbley scoring method \citep{hadler}, developed for screwdriver markings, for same-source identification of bullet striae. We provide methods to identify parameters that minimize the error rates for matching of LEAs using Hamby datasets 44 and 252 measured by NIST and CSAFE. We suggest a remedial algorithm to alleviate the problem of failed tests in the method, increasing both the power of the test and reducing error rates. Type II error rates of the proposed method improve by more than one third (Type I error of 0.05) and are on average about 0.22. This puts the proposed method on similar footing as other single feature matching approaches in the literature. 
---

@@

We then define the pairs for different-shifts as
\begin{equation}\label{eq.diff2}
\left(s_{(j)}^{(1)}, s_{(I-j+1)}^{(1)} \right) \text{ for } j = 
\begin{cases}
1, ..., I & \text{ for even } I \\
1, ..., (I-1)/2, (I-1)/2 + 2, ..., I & \text{ for odd } I
\end{cases},
\end{equation}
i.e.\ for an odd number of same-shift correlations, we skip the middle pair for the different-shift correlations (see \autoref{sketch-diff-2}).
This pairing ensures that the number of different-shift pairings is the same or at most one less than the number of same-shift pairings in all tests.
In the remainder of the paper, we will refer to the algorithm defined by \citet{hadler} as \textbf{(CS1)} and the suggested modified algorithm as \textbf{(CS2)} and compare their performance on the available scans of the Hamby study. The section \ref{appendix:appxfailed} of the Appendix discusses about the various scenarious of failed tests.
\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-diff-2.png}
\caption{\label{sketch-diff-2}Sketch of adjusted different-shift pairings. At most one of the same-shift pairings can not be matched with a different-shift pair under algorithm (CS2). }
\end{figure}

@@

errorsCV <- rbind(errorsCV1, errorsCV2)
ggerrors <- errorsCV %>%
  ggplot(aes(x = wo, y = beta, colour=factor(alpha), linetype=factor(alpha))) +
  geom_point(aes(shape=factor(wv)), size=2.5) +
  geom_smooth(aes(group=alpha), se=FALSE, method="loess", span=1,  size=.7) +
  theme_bw() +
  scale_colour_brewer(expression("Nominal\ntype I error"~alpha),
                      palette="Set2", guide=FALSE) +
  scale_linetype_manual(values = c(2,1,3,4), guide=FALSE)+
  scale_shape_discrete(expression("Size of\nvalidation window "~w[v])) +
  guides(
    colour = guide_legend(override.aes = list(linetype= c(2,1,3,4)))
  )+
  xlab(expression("Window size for optimization "~w[o])) +
  ylab("Type II error rate") +
  facet_wrap(~method, labeller="label_both")

@@

The results from Figures \ref{fig:roc} and \ref{fig:eer} are summarized in numbers in \autoref{tab:aucs}. Equal error rates (EER), rates for false positives (FPR) and false negatives (FNR) are shown side by side with the area under the curve (AUR) for both methods for a set of different optimization windows $w_o$ and a validation window $w_v$ of 30 pixels. The rate of false positive same-source identifications is equal to the statistical type I error rate, which is set to $\alpha = 0.05$ for this example. The rate of false negatives are missed same-source markings. This rate is also known as the type II error rate. A detailed plot on the type II error rates for CS1 and CS2 can be found in the section \ref{appendix:appxtype2} of the Appendix.
Area under the curve (AUC) is shown with confidence intervals as given by \citet{delong}. CS2 significantly outperforms CS1 with respect to its predictive power in most situations.

@@

cvcs %>% filter(!wo %in% c(130, 140)) %>%
  ggplot(aes(d = as.numeric(match), m = 1-p_value)) + 
  geom_roc(labels=FALSE, n.cuts=500, pointsize=0, size=0.6, 
           aes(colour = method.wo, linetype= method.wo)) + 
  facet_wrap(~wv, labeller = label_bquote(w[v] == .(wv))) +
    scale_linetype_manual(values=c(2,2,2,2,2,1,1,1,1,1), guide = FALSE) +
  scale_alpha_discrete(range=c(0.1, 0.6)) +
  scale_colour_manual(expression("Method /\nOptimization\nwindow size"~w[o]), values=cols) +
  theme_bw() +
  xlab("False Positives Rate (1 - Specificity)") +
  ylab("True Positives Rate (Sensitivity)") +
  guides(
   colour = guide_legend(override.aes = list(linetype = c(2,2,2,2,2,1,1,1,1,1)), ncol=2)
  ) +
  geom_abline(slope=-1, intercept = 1, linetype = 3, 
              colour="grey30", size = 0.25)
